### 2016-04-22 manuscript-body.tex
Replace "I" with "we" for joint authorship

I argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have.

@@

Identifying and removing barriers to student learning seems like a good
thing. So, it should follow that cataloging student misconceptions and
developing remedies for them should also be a good thing. But, the
logical implication isn't that clean. In some cases students display
productive, useful knowledge that's either ignored or outright
criticized by researchers. Aligning students toward canonical knowledge
makes sense, but doing so at the expense of---or in direct contradiction
to---useful ways of knowing seems undesirable at best. Next, I expand on
two examples from misconceptions research in programming. Specifically,
I show how and why I think a misconceptions focus in programming casts
aside students' useful intuitions and understandings.

The first example comes from \cite{kaczmarczyk_identifying_2010}. Part of that
study involved giving students snippets of Java code and asking students
to diagram (or pseudo-code) how the information would be stored in
memory. Below I have reproduced the code for Problem 2:

@@

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As I show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, I would argue
there are conceptual ways a \emph{programmer} may need to think about
interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that even Sherin \cite{sherin_how_2001} doesn't
enumerate.One of the most obvious, for example, is the conceptual movement from seeing
  $\Box=︎ \Box +\Box$ as a statement of equality (typical in a math class) to seeing it as the assignment of a
  sum to a variable (typical in programming). To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

@@

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge as a problem. I begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

@@

Let's begin with two motivating examples. My aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. In Table 2 below, I imagine five
different ways of writing a programming statement. Each of the five ways
is semantically equivalent to the others (or near enough for explanatory
purposes). And, in each example I add a layer of specificity to the
syntax. I also propose a corresponding interpretation of how I might
apply interpret and think about a given statement.\footnote{I assume for
  simplicity's sake that all variables and functions in my code samples
  evaluate to floating-point numbers that can be combined under
  addition. I also assume, crucially, that all statements are in the
  same hypothetical language and that the variable names connote nothing
  to the computational interpreter.}

@@

I stress that the third column is about \emph{how I might think about}
each programming statement\emph{.} By no means am I making normative
claims about how one \emph{ought} to think about it, or whether the
statement actually does what its names might suggest it does. Rather,
``How I might think about it'' reflects the kind of local meaning or
interpretation I might attach to such a statement when I work with it,
given my understanding of its role and context.

Consider specifically statements 1 and 4. Statement 1 could very well be
an incrementer in some kind of iterative code. If I had to debug code
that employs statement 1, what I might do is exploit my knowledge of
what d is (does it hold a constant value? Is it 1?) and try inserting
intermediate print statements into the iterative code. Doing so, I can
inspect textually how values change with each iteration. In statement
4's case, it could very well be that the position-update code animates
images on a screen. If there's a problem with that code, one of the
easiest ways I might notice is that the acceleration seems off in the
graphics. Consequently, my debugging might call upon knowledge I have
from kinematics. I might try inserting code that draws a dot at the
object's on-screen position with each iteration, leaving a trail I can
visually inspect. I could then look at the path of the object's
trajectory and the spacing patterns between successive dots as a
first-pass test of whether my code achieves the motion I want.

To be clear, it's not just that graphics might improve my efficiency in
debugging a statement. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets me
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in my code. If I view the assignment statement as saying
something about the motion of an object (cf., Hammer, 1994, p. 165), a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if I focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what I know about physics in order to
reason about the code.

My second example concerns statements that are syntactically-similar,
rather than semantically-similar. I proposed that the statements in
Table 2 all came from the same language. But, another phenomenon comes
into play when different languages use the same symbology for
conceptually different operations. Table 3 shows examples of what look
like semantically-equivalent operations, but in fact are not.

@@

% \begin{supertabular}[]{@{}llll@{}}
% \toprule
% \textbf{Statement } & Programming Syntax & Language & How I think about
% it\tabularnewline
% \midrule
% \endhead
% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
% the string w\tabularnewline
% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
% layer of points onto a plot p\tabularnewline
% \bottomrule
% \end{supertabular}

@@

Given these motivating examples, it seems sensible to think there's
utility in a programmer having different conceptual metaphors available
to think about and work with code. Example 1 shows that choosing to
apply knowledge from physics to a piece of code can change the cognitive
nature of debugging. There, debugging a position-update statement
becomes, in part, reasoning kinematically about the properties of motion
trails.\footnote{For comparison, consider the argument that ringing an
  aircraft speedometer with physical markers changes the nature of
  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
2 shows that across languages, programmers might have to deploy
different conceptual metaphors to reason about statements in a
locally-consistent way. Knowing that plots in ggplot2 can be composed
layer-by-layer with reassignment is crucial if you're trying to write or
understand code that creates statistical graphics. But, I would argue
that thinking about $\Box = \Box + \Box$ as ``compose new layer onto plot'' can and
does appeal to different kinds of knowledge when compared to thinking
about $\Box = \Box + \Box$ as ``include this addend in the sum,'' which itself can
and does appeal to different kinds of concepts when compared to thinking
about $\Box = \Box + \Box$ as ``increment the counter.''

Stepping back, I can build the following argument

>>>>

we argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have.

@@

Identifying and removing barriers to student learning seems like a good
thing. So, it should follow that cataloging student misconceptions and
developing remedies for them should also be a good thing. But, the
logical implication isn't that clean. In some cases students display
productive, useful knowledge that's either ignored or outright
criticized by researchers. Aligning students toward canonical knowledge
makes sense, but doing so at the expense of---or in direct contradiction
to---useful ways of knowing seems undesirable at best. Next, we expand on
two examples from misconceptions research in programming. Specifically,
we show how and why we think a misconceptions focus in programming casts
aside students' useful intuitions and understandings.

The first example comes from \cite{kaczmarczyk_identifying_2010}. Part of that
study involved giving students snippets of Java code and asking students
to diagram (or pseudo-code) how the information would be stored in
memory. Below we have reproduced the code for Problem 2:

@@

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As we show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, we would argue
there are conceptual ways a \emph{programmer} may need to think about
interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that even Sherin \cite{sherin_how_2001} doesn't
enumerate.One of the most obvious, for example, is the conceptual movement from seeing
  $\Box=︎ \Box +\Box$ as a statement of equality (typical in a math class) to seeing it as the assignment of a
  sum to a variable (typical in programming). To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

@@

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge as a problem. we begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

@@

Let's begin with two motivating examples. My aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. In Table 2 below, we imagine five
different ways of writing a programming statement. Each of the five ways
is semantically equivalent to the others (or near enough for explanatory
purposes). And, in each example we add a layer of specificity to the
syntax. we also propose a corresponding interpretation of how we might
apply interpret and think about a given statement.\footnote{we assume for
  simplicity's sake that all variables and functions in my code samples
  evaluate to floating-point numbers that can be combined under
  addition. we also assume, crucially, that all statements are in the
  same hypothetical language and that the variable names connote nothing
  to the computational interpreter.}

@@

we stress that the third column is about \emph{how we might think about}
each programming statement\emph{.} By no means am we making normative
claims about how one \emph{ought} to think about it, or whether the
statement actually does what its names might suggest it does. Rather,
``How we might think about it'' reflects the kind of local meaning or
interpretation we might attach to such a statement when we work with it,
given my understanding of its role and context.

Consider specifically statements 1 and 4. Statement 1 could very well be
an incrementer in some kind of iterative code. If we had to debug code
that employs statement 1, what we might do is exploit my knowledge of
what d is (does it hold a constant value? Is it 1?) and try inserting
intermediate print statements into the iterative code. Doing so, we can
inspect textually how values change with each iteration. In statement
4's case, it could very well be that the position-update code animates
images on a screen. If there's a problem with that code, one of the
easiest ways we might notice is that the acceleration seems off in the
graphics. Consequently, my debugging might call upon knowledge we have
from kinematics. we might try inserting code that draws a dot at the
object's on-screen position with each iteration, leaving a trail we can
visually inspect. we could then look at the path of the object's
trajectory and the spacing patterns between successive dots as a
first-pass test of whether my code achieves the motion we want.

To be clear, it's not just that graphics might improve my efficiency in
debugging a statement. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets me
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in my code. If we view the assignment statement as saying
something about the motion of an object (cf., Hammer, 1994, p. 165), a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if we focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what we know about physics in order to
reason about the code.

My second example concerns statements that are syntactically-similar,
rather than semantically-similar. we proposed that the statements in
Table 2 all came from the same language. But, another phenomenon comes
into play when different languages use the same symbology for
conceptually different operations. Table 3 shows examples of what look
like semantically-equivalent operations, but in fact are not.

@@

% \begin{supertabular}[]{@{}llll@{}}
% \toprule
% \textbf{Statement } & Programming Syntax & Language & How we think about
% it\tabularnewline
% \midrule
% \endhead
% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
% the string w\tabularnewline
% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
% layer of points onto a plot p\tabularnewline
% \bottomrule
% \end{supertabular}

@@

Given these motivating examples, it seems sensible to think there's
utility in a programmer having different conceptual metaphors available
to think about and work with code. Example 1 shows that choosing to
apply knowledge from physics to a piece of code can change the cognitive
nature of debugging. There, debugging a position-update statement
becomes, in part, reasoning kinematically about the properties of motion
trails.\footnote{For comparison, consider the argument that ringing an
  aircraft speedometer with physical markers changes the nature of
  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
2 shows that across languages, programmers might have to deploy
different conceptual metaphors to reason about statements in a
locally-consistent way. Knowing that plots in ggplot2 can be composed
layer-by-layer with reassignment is crucial if you're trying to write or
understand code that creates statistical graphics. But, we would argue
that thinking about $\Box = \Box + \Box$ as ``compose new layer onto plot'' can and
does appeal to different kinds of knowledge when compared to thinking
about $\Box = \Box + \Box$ as ``include this addend in the sum,'' which itself can
and does appeal to different kinds of concepts when compared to thinking
about $\Box = \Box + \Box$ as ``increment the counter.''

Stepping back, we can build the following argument
